---
title: "Introduction to MixCat"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to MixCat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

## Overview

MixCat fits **mixture catalytic models** to antibody titer data, jointly estimating:

- The **force of infection (FOI)** via a catalytic model, giving smooth age-specific seroprevalence
- The **titer distributions** of seronegative and seropositive individuals via a two-component Gaussian mixture

Two main Stan models are provided:

| Model | When to use |
|---|---|
| `MixtureCatalytic` | Single population; assumes endemic transmission |
| `Mixture` | Age-specific seroprevalence estimated directly; no catalytic assumption |
| `MixtureCatalyticGrouped` | Multiple populations with separate FOI (e.g. urban/rural) |

---

## Setup

```{r libraries}
library(rstan)
library(bayesplot)
library(tidyverse)

source("../utils.R")

rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```

---

## Data

MixCat expects a data frame with a titer column and an age group column. Here we use
the bundled simulated dataset, which was generated under a catalytic model with
$\lambda = 0.07$, $\mu_0 = 2$, $\mu_1 = 2.5$, $\sigma_0 = 0.7$, $\sigma_1 = 0.9$.

```{r data}
simdata <- readRDS("../data/SimulatedData.RDS")

# Age group bounds
ageL <- c(0, 10, 20, 30, 40, 50, 60, 70, 80)
ageU <- c(9, 19, 29, 39, 49, 59, 69, 79, 90)

df <- simdata$simdf |>
  mutate(age_group_int = as.integer(as.factor(age_group)),
         age_label     = paste0(ageL[age_group_int], "-", ageU[age_group_int]),
         age_label     = factor(age_label, levels = paste0(ageL, "-", ageU)))

head(df)
```

```{r data-plot, fig.cap="Titer distribution by age group"}
ggplot(df, aes(x = age_label, y = titer)) +
  geom_jitter(width = 0.2, alpha = 0.4, color = "steelblue3") +
  geom_boxplot(alpha = 0.3, outlier.shape = NA) +
  theme_bw() +
  xlab("Age group") + ylab("Titer") +
  theme(text = element_text(size = 13),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r titer-hist, fig.cap="Overall titer distribution"}
ggplot(df, aes(x = titer)) +
  geom_histogram(bins = 30, fill = "steelblue3", color = "white", alpha = 0.8) +
  theme_bw() +
  xlab("Titer") + ylab("Count") +
  theme(text = element_text(size = 13))
```

---

## Preparing model inputs

`make_model_data()` constructs the correctly formatted Stan input list from your titer
vector, age group indices, and prior specifications. Priors are specified for the four
mixture parameters in order: $\mu_0$, $\mu_1$, $\sigma_0$, $\sigma_1$.

```{r model-data}
model_data <- make_model_data(
  titer       = df$titer,
  age_group   = df$age_group_int,
  ageL        = ageL,
  ageU        = ageU,
  prior_means = c(2, 3, 1, 1),
  prior_sds   = c(1, 1, 1, 1)
)

# Check sample sizes per age group
tibble(age_group = paste0(ageL, "-", ageU),
       n         = model_data$n_age)
```

---

## Fitting the models

Models are fitted with `rstan::stan()`. The `iter` and `warmup` arguments control total
and warmup iterations per chain; post-warmup draws = `iter - warmup`.

```{r fit, cache=TRUE, results="hide"}
fit_cat <- stan(
  file    = "../StanModels/MixtureCatalytic.stan",
  data    = model_data,
  chains  = 3, cores  = 1,
  iter    = 1000, warmup = 500,
  refresh = 10
)

fit_mix <- stan(
  file    = "../StanModels/Mixture.stan",
  data    = model_data,
  chains  = 3, cores  = 1,
  iter    = 1000, warmup = 500,
  refresh = 10
)
```

---

## Checking convergence

```{r convergence, fig.cap="Trace plots — MixtureCatalytic"}
color_scheme_set("mix-blue-red")
mcmc_trace(fit_cat, regex_pars = c("lambda", "mu", "sd"))
```

```{r rhat, fig.cap="R-hat values — all parameters should be near 1"}
mcmc_rhat(rhat(fit_cat)) + yaxis_text()
```

Posterior draws are extracted with `rstan::extract()`, returning a named list with one
element per parameter.

```{r extract}
draws_cat <- rstan::extract(fit_cat)
draws_mix <- rstan::extract(fit_mix)
```

---

## MixtureCatalytic outputs

### Seroprevalence by age

The catalytic model estimates a single FOI ($\lambda$) and derives a smooth seroprevalence
curve: $\pi(a) = 1 - e^{-\lambda a}$.

```{r seroprev-cat, fig.cap="Estimated seroprevalence with 95% CrI (catalytic model)"}
plot_seroprev(draws_cat, model_data)
```

The underlying estimates are available as a tidy data frame:

```{r seroprev-cat-extract}
extract_seroprev(draws_cat, model_data) |> head()
```

### Overall titer distribution fit

```{r dist-fit-cat, fig.cap="Observed titer histogram with posterior predictive mixture components"}
plot_dist_fit(draws_cat, model_data)
```

### Observed vs predicted mean titer by age group

```{r mean-titer-cat, fig.cap="Observed (grey) and predicted (green) mean titer by age group with 95% CIs/CrIs"}
plot_mean_titer(draws_cat, model_data)
```

### Individual titer predictions

```{r titer-pred-cat, fig.cap="Observed vs posterior median predicted titer per individual"}
plot_titer_pred(draws_cat, model_data)
```

### Probability of seropositivity

```{r prob-seropos-cat, fig.cap="Posterior median P(seropositive) as a function of observed titer"}
plot_prob_seropos(draws_cat, model_data)
```

---

## Mixture model outputs

The non-catalytic `Mixture` model estimates seroprevalence independently per age group,
without assuming endemic transmission. All titer diagnostic plots are identical to above
since the mixture components are shared — only the seroprevalence output differs.

### Age-specific seroprevalence

```{r seroprev-mix, fig.cap="Estimated seroprevalence by age group with 95% CrI (mixture model)"}
plot_seroprev_mix(draws_mix, model_data)
```

```{r seroprev-mix-extract}
extract_seroprev_mix(draws_mix, model_data)
```

### Distribution fit

```{r dist-fit-mix, fig.cap="Distribution fit (mixture model)"}
plot_dist_fit(draws_mix, model_data)
```

---

## Working with extract outputs

Every `plot_*` function has a corresponding `extract_*` returning a tidy data frame for
further analysis or custom plotting.

```{r custom-plot, fig.cap="Custom seroprevalence plot comparing both models"}
seroprev_cat <- extract_seroprev(draws_cat, model_data) |>
  mutate(model = "Catalytic")

seroprev_mix <- extract_seroprev_mix(draws_mix, model_data) |>
  rename(age = age_mid, seroprev = seroprev, criL = criL, criU = criU) |>
  mutate(model = "Mixture")

# Overlay catalytic curve with mixture age-group estimates
ggplot() +
  geom_ribbon(data = seroprev_cat,
              aes(x = age, ymin = criL, ymax = criU),
              fill = "indianred2", alpha = 0.3) +
  geom_line(data = seroprev_cat,
            aes(x = age, y = seroprev),
            color = "indianred2", linewidth = 1) +
  geom_linerange(data = seroprev_mix,
                 aes(x = age, ymin = criL, ymax = criU),
                 color = "steelblue3", linewidth = 0.8) +
  geom_point(data = seroprev_mix,
             aes(x = age, y = seroprev),
             color = "steelblue3", size = 3) +
  annotate("text", x = c(70, 70), y = c(0.85, 0.75),
           label = c("Catalytic", "Mixture"),
           color = c("indianred2", "steelblue3"), size = 5) +
  theme_bw() +
  xlab("Age") + ylab("Seroprevalence") +
  ylim(0, 1) +
  theme(text = element_text(size = 14))
```

---

## True parameter values

Since this is simulated data, we can check estimates against the known truth.

```{r true-params}
simdata$truepars
```

```{r param-comparison}
tibble(
  parameter  = c("lambda", "mu0", "mu1", "sd0", "sd1"),
  true_value = with(simdata$truepars, c(lam, mu0, mu1, sd0, sd1)),
  post_median = c(
    median(draws_cat$lambda),
    median(draws_cat$mu0),
    median(draws_cat$mu1),
    median(draws_cat$sd0),
    median(draws_cat$sd1)
  ),
  post_lo = c(
    quantile(draws_cat$lambda, 0.025),
    quantile(draws_cat$mu0,    0.025),
    quantile(draws_cat$mu1,    0.025),
    quantile(draws_cat$sd0,    0.025),
    quantile(draws_cat$sd1,    0.025)
  ),
  post_hi = c(
    quantile(draws_cat$lambda, 0.975),
    quantile(draws_cat$mu0,    0.975),
    quantile(draws_cat$mu1,    0.975),
    quantile(draws_cat$sd0,    0.975),
    quantile(draws_cat$sd1,    0.975)
  )
) |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))
```

---

## Session info

```{r session-info}
sessionInfo()
```
