---
title: "Introduction to MixCat"
output: github_document
vignette: >
  %\VignetteIndexEntry{Introduction to MixCat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

## Overview

MixCat fits **mixture** and **mixture-catalytic** models to antibody titer data. Both models represent the population titer distribution as a two-component Gaussian mixture, where the two components represent seronegative ($\mu_0$, $\sigma_0$) and seropositive ($\mu_0 + \mu_1$, $\sigma_1$) individuals. Age-specific seroprevalence is estimated either directly from antibody titer data  using a mixture model (`Mixture.stan`) or jointly with a force of infection (FOI) parameter via a mixture-catalytic model (`MixtureCatalytic.stan`). Both models support stratified estimates by sub-populations (e.g. urban/rural) via an optional `group` argument.


#### When to use each model

In general, these models should only be used when the population antibody titer distribution has a bimodal pattern (i.e. some separation of antibody responses between positive and negative individuals) OR if the approximate range of antibody responses for positive or negative individuals is known from control samples, which can be used to inform model priors.

**`Mixture`** makes no assumption about pathogen transmission dynamics and can therefore be used in all scenarios. Age-specific seroprevalence estimates from this model can give an agnostic indication of past transmission dynamics (e.g. endemic or epidemic) for pathogens that induce life-long serological responses.

**`MixtureCatalytic`** can be used when endemic transmission is a reasonable assumption, for example when a trend of increasing seroprevalence by age was observed from the **`Mixture`** model.

Both models support the estimation of age-specific seroprevalence / FOI by sub-populations (e.g. urban/rural locations, sex, socioeconomic status).

#### Assumptions

-   Both models assume the population to be composed of seronegative and seropositive individuals. Model estimates may be unreliable if true seroprevalence is close to 0% or 100%.

-   Both models assume no antibody waning or seroreversion. Seroprevalence will be underestimated for pathogens/antigens where antibodies wane over short timescales.

-   **`MixtureCatalytic`** assumes a constant endemic force of infection (FOI), i.e. no differences in the risk of infection by age or over time. For pathogens with seasonal or annual fluctuations in transmission intensity, the FOI parameter will represent a long-term average annual estimate of this transmission intensity.


---

## Setup

Clone the repository and source the helper functions:


```{r libraries}
library(rstan)
library(bayesplot)
library(tidyverse)

source("../utils.R")

rstan_options(auto_write = TRUE)            # cache compiled Stan models to disk
options(mc.cores = parallel::detectCores()) # run chains in parallel
```

---

## Data Preparation

MixCat inputs:

- **`titer`**: a numeric vector of antibody titer values
- **`age_group`**: an ordered factor assigning each individual to an age group
- **`group`** *(optional)*: a factor identifying sub-population groups (e.g. location) for stratified estimates

Here we use a simulated dataset of 400 individuals with age group and location (urban vs rural) information. To prepare the data for model fitting, an ordered integer age group variable is required such that the youngest age group = 1 and the oldest age group = total number of age groups. It is important to check that you have sufficient sample size in each age group before model fitting. Wider age bins can be used when sample sizes are low.

```{r data}
df <- load_example_data("../data/SimulatedData.RDS")
head(df)

# Check sample sizes within age groups
table(df$age_group)

# Specify age group bounds — must match the levels of the age_group factor in the data
ageL <- c(0, 10, 20, 30, 40, 50, 60, 70, 80)
ageU <- c(9, 19, 29, 39, 49, 59, 69, 79, 90)
```

Before fitting, it is worth visualising the antibody titer distribution to check whether the data are appropriate for a Gaussian mixture model. The data are assumed to be composed of two Gaussian distributions — for antibody measurements on large scales, titer measurements should be log-transformed before model fitting. This simulated data was generated on a log titer scale and we can see a rough bimodal pattern, plotted below.

```{r titer-hist}
ggplot(df, aes(x = titer)) + theme_bw() +
  geom_histogram(bins = 40, fill = "grey", col = "white", position = "identity") +
  xlab("log titer") + ylab("Count") + theme(text = element_text(size = 13))
```

---

## Model inputs

`make_model_data()` assembles the Stan model input list. The `prior_means` and `prior_sds` arguments specify Normal priors for the four mixture parameters ($\mu_0$, $\mu_1$, $\sigma_0$, $\sigma_1$) in that order. Choose priors that are consistent with the scale of your titer data — the values used below reflect log-scale titers centered near 2–5. Note that $\mu_1$ is the mean rise in antibody titers from seronegative to seropositive, such that the mean of the seropositive distribution is defined as $\mu_0 + \mu_1$.

Here we will create two versions of the model input lists, one for estimating overall population parameters and another for stratified estimation by location (using the `group` argument).

```{r model-data}
# model inputs for overall population model
model_data <- make_model_data(
  titer       = df$titer, # antibody titer measurements
  age_group   = df$age_group_int, # ordered age group integer variable
  ageL        = ageL, # lower age group bounds
  ageU        = ageU, # upper age group bounds
  prior_means = c(2, 3, 1, 1), # prior means (Normal distribution)
  prior_sds   = c(1, 1, 1, 1) # prior SDs (Normal distribution)
)

# model inputs for estimation stratified by location
model_data_loc <- make_model_data(
  titer       = df$titer, # antibody titer measurements
  age_group   = df$age_group_int, # ordered age group integer variable
  ageL        = ageL, # lower age group bounds
  ageU        = ageU, # upper age group bounds
  group       = df$location_int, # location integer variable
  prior_means = c(2, 3, 1, 1), # prior means (Normal distribution)
  prior_sds   = c(1, 1, 1, 1) # prior SDs (Normal distribution)
)
```

---

## Mixture model
We will first fit a mixture model, where age-specific seroprevalence is directly inferred from the antibody titer distributions with no assumptions about transmission dynamics.

### Fitting

Models are fitted with `rstan::stan()`. The settings below (500 warmup, 1,000 sampling iterations across 3 chains = 3000 posterior draws) are sufficient for most datasets. Increase `iter` if R-hat values are elevated or trace plots show poor mixing.

```{r fit-mix, cache=TRUE, results="hide"}
fit_mix <- stan(
  file    = "../StanModels/Mixture.stan",
  data    = model_data,
  chains  = 3, cores  = 1,
  iter    = 1500, warmup = 500,
  refresh = 100
)
```

### Checking convergence

Trace plots and R-hat values should be checked to assess model convergence. R-hat values should be $\leq 1.01$ — values above 1.05 indicate poor convergence and model estimates should not be interpreted.

```{r convergence-mix}
color_scheme_set("mix-blue-red")
mcmc_trace(fit_mix, regex_pars = c("sero", "mu", "sd"))
```

```{r rhat-mix}
rhats <- rhat(fit_mix)
mcmc_rhat(rhats[grepl("^(sero|mu|sd)", names(rhats))]) + yaxis_text() + theme_bw()
```


### Model estimates

Overall population seroprevalence estimates can be extracted with the `extract_sero` function. Here the model estimates 60% (95% credible interval: 54-65%) population seroprevalence. Mixture parameter estimates ($\mu_0$, $\mu_1$, $\sigma_0$, $\sigma_1$) can be extracted with `extract_mixture_params`. The function `extract_sero_age_mix` will return a dataframe of age-specific seroprevalence estimates from the `Mixture` model and the `plot_sero_age_mix` function will create a simple plot of these estimates. If model estimates were stratified by other groups, e.g. location, the `group_labels` argument can be used to specify labels. 

```{r extract-mix}
# extract model draws
draws_mix <- rstan::extract(fit_mix)

# extract seroprevalence estimates
extract_sero(draws = draws_mix, model_data = model_data, group_labels = NULL)

# extract mixture parameters
extract_mixture_params(draws = draws_mix)
```

```{r seroprev-mix-extract}
# get age-specific seroprevalence estimates from Mixture model
extract_sero_age_mix(draws = draws_mix, model_data = model_data, group_labels = NULL)

# plot age-specific seroprevalence estimates from Mixture model
plot_sero_age_mix(draws = draws_mix, model_data = model_data, group_labels = NULL)

```

### Assess model fit

The `plot_dist_fit` function can be used to visualize the model fit to the observed data. Grey bars show the observed antibody titer data while coloured lines (median estimates) and ribbons (95% credible interval estimates) show the model reconstructed antibody titer distributions. For custom plotting, the `extract_dist_fit` function can be used to obtain the data and estimates underlying this plot. Another way of assessing model fit is comparing observed and estimated mean titers by age group - this can be done using the `extract_mean_titer` and `plot_mean_titer` functions.


```{r dist-fit-mix}
# plot observed vs model reconstructed titer distributions
plot_dist_fit(draws_mix, model_data)

# plot observed vs model estimated mean titers by age groups
plot_mean_titer(draws = draws_mix, model_data = model_data, group_labels = NULL)

```

### Probability of seropositivity

The posterior probability of being seropositive given observed titers can be extracted and plotted using the `extract_prob_seropos` and `plot_prob_seropos` functions. This can be a useful metric for understanding where uncertainty is highest in individual serostatus classification.

```{r prob-seropos}
plot_prob_seropos(draws_mix, model_data, group_labels = NULL)
```

---

## MixtureCatalytic model

The **`MixtureCatalytic`** model jointly estimates a force of infection parameter  ($\lambda$) and age-specific seroprevalence from the mixture components. Age-specific seroprevalence, $\pi(a)$, is derived from the FOI by $\pi(a) = 1 - e^{-\lambda a}$. This model assumes endemic transmission, enforcing a trend of increasing seroprevalence by age.

### Fitting

Models are fitted with `rstan::stan()`. The settings below (500 warmup, 1000 sampling iterations across 3 chains = 3000 posterior draws) are sufficient for most datasets. Increase `iter` if R-hat values are elevated or trace plots show poor mixing.

Here, we will fit the **`MixtureCatalytic`** model allowing for stratified estimates by location (using the `model_data_loc` object created before).
```{r fit-cat, cache=TRUE, results="hide"}

fit_cat <- stan(
  file    = "../StanModels/MixtureCatalytic.stan",
  data    = model_data_loc,
  chains  = 3, cores  = 1,
  iter    = 1500, warmup = 500,
  refresh = 100
)
```

### Checking convergence

As before, we will look at trace plots and R-hat values to check model convergence. R-hat values should be $\leq 1.01$ — values above 1.05 indicate poor convergence and model estimates should not be interpreted.
```{r convergence-cat}
color_scheme_set("mix-blue-red")
mcmc_trace(fit_cat, regex_pars = c("lambda", "mu", "sd"))

```

```{r rhat-cat}
rhats_cat <- rhat(fit_cat)
mcmc_rhat(rhats_cat[grepl("^(lambda|mu|sd)", names(rhats_cat))]) + yaxis_text() + theme_bw()

```

### Model estimates

Population seroprevalence and FOI estimates can be extracted with the `extract_sero` and `extract_foi()` functions. Here we estimate a lower FOI in rural locations, 0.02 (95%CrI: 0.01-0.02), compared to 0.06 (95%CrI: 0.04-0.08) in urban locations. This can be interpreted as 2% (95%CrI: 1-2%) and 6% (95%CrI: 4-8%) of the susceptible population become infected each year in rural and urban locations respectively. Mixture parameter estimates ($\mu_0$, $\mu_1$, $\sigma_0$, $\sigma_1$) can be extracted with `extract_mixture_params`.

```{r cat-estimates}
# extract model draws
draws_cat <- rstan::extract(fit_cat)

# extract FOI estimates by location
extract_foi(draws_cat, model_data_loc, group_labels = c("Rural", "Urban"))

# extract population seroprevalence estimates by location
extract_sero(draws_cat, model_data_loc, group_labels = c("Rural", "Urban"))

# extract mixture parameter estimates (shared across groups)
extract_mixture_params(draws_cat)
```



```{r seroprev-cat}
# extract age-specific seroprevalence estimates by location
extract_sero_age_cat(draws_cat, model_data_loc, group_labels = c("Rural", "Urban")) |> head()

# plot age-specific seroprevalence estimates
plot_sero_age_cat(draws_cat, model_data_loc, group_labels = c("Rural", "Urban"))
```


### Assess model fit

The `plot_dist_fit` function can be used to visualize the model fit to the observed data. Grey bars show the observed antibody titer data while coloured lines (median estimates) and ribbons (95% credible interval estimates) show the model reconstructed antibody titer distributions. For custom plotting, the `extract_dist_fit` function can be used to obtain the data and estimates underlying this plot. Another way of assessing model fit is comparing observed and estimated mean titers by age group - this can be done using the `extract_mean_titer` and `plot_mean_titer` functions.


```{r dist-fit-cat}
# plot observed vs model reconstructed titer distributions
plot_dist_fit(draws_cat, model_data_loc)

# plot observed vs model estimated mean titers by age groups
plot_mean_titer(draws = draws_cat, model_data = model_data_loc, group_labels = c("Rural", "Urban"))

```


### Probability of seropositivity

The posterior probability of being seropositive given observed titers can be extracted and plotted using the `extract_prob_seropos` and `plot_prob_seropos` functions. This can be a useful metric for understanding where uncertainty is highest in individual serostatus classification.

```{r prob-seropos-cat}
plot_prob_seropos(draws_cat, model_data_loc, group_labels = c("Rural", "Urban"))
```

